---
title: "Exercise 1 --- Basic Maps and Graphs in R"
subtitle: "BSTA 374 · 12 points · Due Thursday, September 10, 2026, 11:59 pm"
author: "Ahmad Tariq · Lehigh University"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "On this page"
    number-sections: false
    code-copy: true
    highlight-style: github
    embed-resources: true
prefer-html: true
execute:
  eval: true
  echo: true
---

## What this exercise is

Your first graded work in this course. It covers the first two weeks: reading
spatial data with `sf`, understanding what a coordinate reference system is and
moving between them, and producing honest choropleth and point maps with both
`ggplot2` and `tmap`.

**No data to download.** Everything runs on the North Carolina SIDS data that
ships inside the `sf` package, so your analysis will reproduce on my machine
exactly as it does on yours --- which is the point.

Each part ends with **starting code**. Copy it, fill in the blanks marked
`______`, and adapt it. It is scaffolding, not a solution: the mechanical lines
are given so that your time goes into the parts that require judgement.

::: {.callout-important}
## Submission
Submit **one file** to Course Site:

- `LASTNAME_FI_Exercise1.qmd` --- the source

Example: `DELMELLE_EM_Exercise1.qmd`. Put your name in the document header.

**Due Thursday, September 10, 2026 at 11:59 pm.** Late work loses 20% per day
unless you arrange something with me in advance.
:::

::: {.callout-warning}
## Reproducibility is graded
Before you submit: restart R (`Ctrl/Cmd + Shift + F10`) and render the document
from a clean session. If it does not run start to finish, it is incomplete ---
regardless of how good the maps are. This is the habit that matters most, and
we start enforcing it now.
:::

```{r}
#| label: setup

library(sf)
library(dplyr)
library(ggplot2)
library(tmap)
# add anything else you use -- and load it here, not halfway down

nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
```

---

## Part A --- Read and describe the data (2 points)

Answer each in **one or two sentences of prose**, outside the code chunks, and
show the code you used to find out.

1. How many features does `nc` contain, and what does one row represent?
```{r}
nrow(nc)
names(nc)
nc
```
  The nc dataset contains 100 features, and each row represents one county in North Carolina. 
  
2. What geometry type is stored? Why is it that type rather than `POLYGON`?
```{r}
table(st_geometry_type(nc))
```
  The geometry type stored is MULTIPOLYGON. It is used instead of POLYGON because some North Carolina counties, especially along the coast, have separate disconnected pieces such as barrier islands that are still part of the same county. 

3. What coordinate reference system is the data in? Give its name, say whether it is geographic or projected, and say how you found out.

```{r}
st_crs(nc)               
st_crs(nc)$input         
st_is_longlat(nc)        
```
  The data use the NAD27 coordinate reference system, which is a geographic CRS using longitude and latitude in degrees. I found this using st_crs(nc), st_crs(nc)$input, and st_is_longlat(nc), which returned TRUE. 

4. In your own words, what do `BIR74` and `SID74` measure? Why does having both matter more than having either one alone?
```{r}
?nc
```
BIR74 measures the number of live births from 1974-1978, while SID74 measures the number of sudden infant deaths during the same period. Having both is important because the number of births gives context for the number of deaths and allows us to calculate a SIDS rate, making comparisons between counties more meaningful. 

## Part B --- Coordinate reference systems (2 points)

1. Transform `nc` to **EPSG:32119** (NAD83 / North Carolina State Plane, in metres) and store it as a **new object**. Do not overwrite `nc` --- you need both.
```{r}
nc_proj <- st_transform(nc, 32119)
```
I transformed nc to EPSG:32119 and stored the projected data in a new object called nc_proj.

2. Plot the counties **before and after**, near enough to each other that a reader can compare them. One or two sentences on what visibly changed.
```{r}
p1 <- ggplot(nc)      + geom_sf(fill = "grey90", colour = "white") + theme_void()
p2 <- ggplot(nc_proj) + geom_sf(fill = "grey90", colour = "white") + theme_void()

p1
p2
```
The projected map changes the visible shape of North Carolina compared with the original geographic map. The original map appears more stretched east to west, while the State Plane projection reduces that distortion for North Carolina. 

3. Compute the area of every county in **square kilometres**, using the projected version. Report the five largest.
```{r}
nc_proj <- nc_proj |>
  mutate(area_km2 = as.numeric(units::set_units(st_area(nc_proj), km^2)))

nc_proj |>
  st_drop_geometry() |>
  slice_max(area_km2, n = 5) |>
  select(NAME, area_km2)
```
The five largest counties by area are Columbus (2450.35 km^2), Robeson (2439.24 km^2), Sampson (2437.41 km^2), Bladen (2288.42 km^2), and Wake (2194.26 km^2).

4. Now compute areas from the *unprojected* object as well. Compare the two sets of numbers and explain, in two or three sentences, what `sf` did and why the projected version is the one you would report in a paper.
```{r}
nc_unproj <- nc |>
  mutate(area_km2 = as.numeric(units::set_units(st_area(nc), km^2)))

nc_unproj |>
  st_drop_geometry() |>
  slice_max(area_km2, n = 5) |>
  select(NAME, area_km2)
```
The areas calculated from the unprojected and projected data are very close but not exactly the same, with Columbus County measuring 2450.831 km² unprojected compared with 2450.349 km² projected. For the unprojected data, sf calculated area using the curved Earth surface, while EPSG:32119 represents North Carolina on a projected flat surface in metres. I would report the projected areas because the North Carolina State Plane projection is designed for accurate area and distance measurements within the state.

::: {.callout-note}
Step 4 will not give you an error, and the two sets of numbers will be close but
not identical. That is the interesting part --- the question is which *surface*
each answer refers to.
::: 
---

## Part C --- Choropleth maps (5 points)

This is the largest part of the exercise, and it is as much about judgement as about code.
1. Add a column giving the **SIDS rate per 1,000 live births** for 1974--78.
```{r}
nc <- nc |>
  mutate(rate74 = 1000 * SID74 / BIR74)
```
The rate74 variable is added as a new column that gives the number of SIDS deaths per 1,000 live births for each county from 1974-1978.

2. Make a choropleth of the **raw counts** (`SID74`) and one of the **rate**, using `ggplot2`. Then write three or four sentences explaining why the two maps disagree, and which one you would put in front of a health department, and why.
```{r}
#Raw count:
ggplot(nc) +
  geom_sf(aes(fill = SID74), colour = "white", linewidth = 0.15)+
  scale_fill_viridis_c(option = "magma", direction = -1, name = "SIDS deaths") +
  labs(title = "SIDS deaths, 1974-1978") +
  theme_void()

#SIDS rate:
ggplot(nc) + 
  geom_sf(aes(fill = rate74), colour = "white", linewidth = 0.15) + 
  scale_fill_viridis_c(option = "magma", direction = -1, name = "per 1,000") + 
  labs(title = "SIDS rate, 1974-78") + 
  theme_void()
```
The two maps show different patterns because the raw SIDS counts are influenced by the number of births in each county. Counties with more births may have more SIDS deaths even when the risk of SIDS is not especially high. The rate accounts for the number of live births, making comparisons between counties with different population sizes more meaningful. I would present the rate map to a health department because it better represents differences in SIDS risk across counties. 

3. Rebuild the rate map using **`tmap`**. Same information, other package.
```{r}
tmap_mode("plot")

tm_shape(nc) +
  tm_polygons(
    fill = "rate74",
    fill.scale = tm_scale_continuous(values = "viridis"),
    fill.legend = tm_legend(title = "SIDS per 1,000 births"),
    col = "white",
    lwd = 0.4
  ) +
  tm_title("SIDS rate, 1974-1978") +
  tm_layout(frame = FALSE)
```
The tmap shows the same county level SIDS rate as the ggplot2 map with higher values concentrated in a smaller number of counties and lower values across much of the state. Using a continuous color scale preserves the variation in SIDS rates while presenting the same information using a different mapping package. 

4. Vary the **class intervals** on the rate map: produce it at least three ways --- for instance equal intervals, quantiles, and Jenks natural breaks. Two or three sentences on how much the apparent geography of risk changes. Name the classification you would defend, and say why.
```{r}
tm_shape(nc) +
  tm_polygons(
    fill = "rate74",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5, values = "brewer.yl_or_rd"),
    fill.legend = tm_legend(title = "SIDS per 1,000 births"), 
    col = "white",
    lwd = 0.4
  ) +
  tm_title("Quantiles") +
  tm_layout(frame = FALSE)


tm_shape(nc) +
  tm_polygons(
    fill = "rate74",
    fill.scale = tm_scale_intervals(style = "equal", n = 5, values = "brewer.yl_or_rd"),
    fill.legend = tm_legend(title = "SIDS per 1,000 births"),
    col = "white",
    lwd = 0.4
  ) +
  tm_title("Equal Intervals") +
  tm_layout(frame = FALSE)

tm_shape(nc) +
  tm_polygons(
    fill = "rate74",
    fill.scale = tm_scale_intervals(
      style = "fisher",
      n = 5,
      values = "brewer.yl_or_rd"
    ),
    fill.legend = tm_legend(title = "SIDS per 1,000 births"),
    col = "white",
    lwd = 0.4
  ) +
  tm_title("Natural Breaks") +
  tm_layout(frame = FALSE)
```
   The apparent geography of SIDS risk changes quite a bit depending on the classification method. Equal intervals place most counties in the lower classes and make only a few extreme counties stand out, while quantiles spread counties more evenly across the five classes and natural breaks emphasize groups that occur naturally in the data. I would defend the quantile classification because it makes regional differences easier to compare across the state, although it can place counties with similar rates into different classes.

::: {.callout-tip}
## What earns full marks here
Not the prettiest map --- the map whose choices you can defend. A student who picks quantiles and explains the trade-off in two clear sentences scores higher than one who picks Jenks because it looked nicer.
:::

5. Every map you submit must have a **legend with units**, a **title**, thin boundaries that do not overwhelm the fill, and a **perceptually uniform, colour-blind-safe palette**. State in one sentence which palette you chose and why, including whether dark means high or low.

I used the color blind safe sequential palettes throughout with magma for the ggplot2 maps, viridis for the continuous tmap map, and YlOrRd for the classified maps, where darker colours represent higher values in the magma and YlOrRd maps and darker colours represent lower values in the viridis map. 

::: {.callout-warning}
## A note on `tmap` versions
`tmap` changed its argument names in version 4. If an example you find online
uses `tm_polygons(col = ...)` or `style = ...` directly and yours errors, check
`packageVersion("tmap")` and read the help page rather than guessing. The code
above is written for version 4.
:::

---

## Part D --- A point layer (2 points)

1. Look up the approximate longitude and latitude of **four cities in North Carolina** and build a small table holding the name and the two coordinates. Cite where you got the coordinates.
```{r}
cities <- tibble(
  name = c("Asheville", "Charlotte", "Raleigh", "Greenville"),
  lon  = c(-82.551491, -80.843124, -78.638176, -77.366356),
  lat  = c(35.595058, 35.227085, 35.779591, 35.612659)
)
```
The approximate coordinates for Asheville, Charlotte, Raleigh, and Greenville were obtained from a LatLong.net's finder list [https://www.latlong.net/.]

2. Turn that table into an `sf` object with the correct CRS. State which EPSG code you used and why.
```{r}
cities_sf <- cities |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)
```
I used EPSG:4326 because the city coordinates are longitude and latitude values in degrees using the WGS84 reference system. 

3. Plot the four cities **on top of** your rate choropleth, labelled legibly.
```{r}
ggplot() +
  geom_sf(data = nc, aes(fill = rate74), colour = "white", linewidth = 0.15) +
  geom_sf(data = cities_sf, size = 2.5, colour = "black") +
  geom_sf_text(data = cities_sf, aes(label = name), size = 3, nudge_y = 0.15) +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "per 1,000") +
  labs(title = "SIDS rate and selected North Carolina cities, 1974-1978") + 
  theme_void()
```
  The four selected cities are clearly shown on the top of the SIDS rate choropleth, making it easy to compare their locations with the surrounding county-level rates. The points and labels are placed above the polygon layer so the city locations remain visible without obscuring the overall geographic pattern. 

4. One sentence: what would go wrong if you built the `sf` object with the coordinate columns in the other order?
  If the longitude and latitude were reversed, the cities would be assigned incorrect coordinates and would be plotted in the wrong locations instead of within North Carolina. 
---

## Part E --- Reproducibility and AI use (1 point)

```{r}
sessionInfo()
```
I used ChatGPT to help debug R code and interpret parts of the assignment that I initially had trouble understanding. I wrote out each section of code and the explanations myself while reviewing lecture notes. I also ran the final code to make sure everything ran smoothly and worked correctly.
---

## Rubric

| Part | What is assessed | Points |
|---|---|---|
| **A** | Correct inspection of the object; clear, accurate answers | 2 |
| **B** | Correct transformation and areas; comparison explained | 2 |
| **C** | Both packages working; counts-vs-rates argument; class intervals compared and a defended choice; cartographic standards met | 5 |
| **D** | Point layer built with the right CRS and correctly overlaid | 2 |
| **E** | Renders from a clean session; AI use note present | 1 |
| | **Total** | **12** |

Writing counts for roughly 10% of the mark within each part: the prose should be
in complete sentences and free of obvious errors. Every part asks for
interpretation as well as code --- a submission that is all code and no sentences
cannot score above about 60%.

---

## Getting help

- Office hours, or email me --- with the code and the exact error message, not a
  screenshot of the whole console.
- The `sf` documentation is genuinely good: `vignette("sf1", package = "sf")`.
- Moraga Chapter 2 covers most of Parts A--C.
- If you are stuck for more than about thirty minutes on something that feels
  mechanical, that is a signal to ask rather than to persevere.